---
title: "Analyzing Top European Companies"
author: "Károly Takács"
date: "2025-02-08"
output: html_document
---

# Introduction

When people think about big stocks, they often jump to Big Tech in the United States. But Europe also
boasts heavyweight corporations with a global footprint-such as LVMH, Siemens, Nestlé, ASML, and more.
Understanding their price movements, correlations, and optimal portfolio allocations is crucial for:
  - Risk management: Minimizing potential drawdowns by balancing exposure.
  - Return optimization: Locating that elusive sweet spot of risk vs. reward.
  - Market insights: Seeing how regional events influence these European giants.

## Loading Libraries
```{r load-packages, message=FALSE, warning=FALSE}
library(sf)                 
library(zoo)
library(purrr)              
library(plotly)              
library(ggplot2)            
library(ggrepel)         
library(quantmod)         
library(quadprog)          
library(gganimate)          
library(tidyverse)          
library(lubridate)          
library(tidygeocoder)      
library(rnaturalearth)      
library(PerformanceAnalytics)
```


## Map of Headquarters:

```{r }
# Prepare European country shapes
europe_shapes <- ne_countries(scale = "medium", continent = "Europe", returnclass = "sf")

# Define the mapping of companies, cities, and countries
company_location <- tibble(
  company = c("LVMH", "Siemens", "Nestlé", "ASML", "Roche", "TotalEnergies", "SAP", "Volkswagen", "BNP Paribas", "Airbus"),
  city = c("Paris", "Munich", "Vevey", "Veldhoven", "Basel", "Courbevoie", "Walldorf", "Wolfsburg", "Paris", "Blagnac"),
  country = c("France", "Germany", "Switzerland", "Netherlands", "Switzerland", "France", "Germany", "Germany", "France", "France")
)

# Geocode
hq_locations <- company_location %>%
  mutate(full_address = paste(city, country, sep = ", ")) %>%
  geocode(address = full_address, method = "osm")

# Count the number of companies in each country
country_counts <- company_location %>%
  count(country, name = "company_count")

# Merge the country counts with European country shapes
map_data <- europe_shapes %>%
  filter(admin %in% country_counts$country) %>%
  left_join(country_counts, by = c("admin" = "country"))

# Plot the map
ggplot(europe_shapes) +
  geom_sf(fill = "gray95", color = "white") +
  geom_sf(data = map_data, aes(fill = company_count), color = "black", alpha = 0.8) +
  geom_point(data = hq_locations, aes(x = long, y = lat), color = "red", size = 1.5) +
  geom_label_repel(data = hq_locations, aes(x = long, y = lat, label = company),
                   size = 3, fontface = "bold") +
  scale_fill_gradient(low = "lightblue", high = "darkblue", name = "Number of Companies by Country") +
  labs(title = "Top European Companies HQ highlight", x = NULL, y = NULL) +
  coord_sf(xlim = c(-10, 20), ylim = c(40, 55), expand = FALSE) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## Stock Price Data from Yahoo Finance with quantmod:

```{r }
# Fetch stock prices for the top 10 European companies from Yahoo Finance
getSymbols(c("MC.PA", "SIE.DE", "NESN.SW", "ASML.AS", "ROG.SW", 
             "TTE.PA", "SAP.DE", "VOW3.DE", "BNP.PA", "AIR.PA"), 
           src = "yahoo", from = "2020-01-01", to = Sys.Date())

# Convert each stock to a data frame and rename columns
stock_data_list <- list(
  "MC.PA"    = fortify.zoo(Cl(`MC.PA`))    %>% rename(date = Index, lvmh_close       = `MC.PA.Close`),
  "SIE.DE"   = fortify.zoo(Cl(`SIE.DE`))   %>% rename(date = Index, siemens_close    = `SIE.DE.Close`),
  "NESN.SW"  = fortify.zoo(Cl(`NESN.SW`))  %>% rename(date = Index, nestle_close     = `NESN.SW.Close`),
  "ASML.AS"  = fortify.zoo(Cl(`ASML.AS`))  %>% rename(date = Index, asml_close       = `ASML.AS.Close`),
  "ROG.SW"   = fortify.zoo(Cl(`ROG.SW`))   %>% rename(date = Index, roche_close      = `ROG.SW.Close`),
  "TTE.PA"   = fortify.zoo(Cl(`TTE.PA`))   %>% rename(date = Index, total_close      = `TTE.PA.Close`),
  "SAP.DE"   = fortify.zoo(Cl(`SAP.DE`))   %>% rename(date = Index, sap_close        = `SAP.DE.Close`),
  "VOW3.DE"  = fortify.zoo(Cl(`VOW3.DE`))  %>% rename(date = Index, volkswagen_close = `VOW3.DE.Close`),
  "BNP.PA"   = fortify.zoo(Cl(`BNP.PA`))   %>% rename(date = Index, bnp_close        = `BNP.PA.Close`),
  "AIR.PA"   = fortify.zoo(Cl(`AIR.PA`))   %>% rename(date = Index, airbus_close     = `AIR.PA.Close`)
)

# Merge all stock data into a single data frame
europe_stock_data <- reduce(stock_data_list, full_join, by = "date") %>%
  mutate(year = year(date), month = month(date))

# Fill missing values using Last Observation Carried Forward (na.locf)
europe_stock_data <- europe_stock_data %>%
  mutate(across(ends_with("_close"), ~ na.locf(.x, na.rm = FALSE)))
```

## Interactive Line Plot of Closing Prices:

```{r }
# Transform data for interactive plot
europe_stock_long <- europe_stock_data %>%
  pivot_longer(cols = ends_with("_close"), names_to = "company", values_to = "price") %>%
  filter(!is.na(price)) %>%
  mutate(company = toupper(gsub("_close", "", company)))

# Create an interactive stock price plot
plot_ly(europe_stock_long, 
        x = ~date, y = ~price, color = ~company, 
        colors = c("blue", "orange", "green", "red", "purple", 
                   "brown", "pink", "gray", "yellow", "cyan"), 
        type = 'scatter', mode = 'lines') %>%
  layout(
    title = "Stock Price Trends: Top 10 European Companies",
    xaxis = list(title = "Date"),
    yaxis = list(title = "Price (EUR)"),
    legend = list(title = list(text = "Company"))
  )
```


## Rolling Volatility 30 day:

```{r }
# Calculate 30-day rolling volatility
volatility_df <- europe_stock_data %>%
  mutate(across(ends_with("_close"), 
                ~ rollapply(.x, 30, sd, fill = NA, align = "right")))

# Transform to long format
volatility_long <- volatility_df %>%
  pivot_longer(cols = ends_with("_close"), names_to = "company", values_to = "volatility") %>%
  mutate(company = toupper(gsub("_close", "", company)))

# Interactive rolling volatility plot
plot_ly(volatility_long, 
        x = ~date, y = ~volatility, color = ~company, 
        colors = c("blue", "orange", "green", "red", "purple", 
                   "brown", "pink", "gray", "yellow", "cyan"), 
        type = 'scatter', mode = 'lines') %>%
  layout(
    title = "30-Day Rolling Volatility",
    xaxis = list(title = "Date"),
    yaxis = list(title = "Volatility (%)"),
    legend = list(title = list(text = "Company"))
  )
```


## Correlation Heatmap:

```{r }
# Filter for the most recent month
most_recent_correlation <- europe_stock_data %>%
  mutate(month_year = floor_date(date, "month")) %>%
  filter(month_year == max(month_year)) %>%
  reframe(across(ends_with("_close"), ~ (./lag(.) - 1) * 100, .names = "{.col}_return")) %>%
  select(ends_with("_return")) %>%
  select(where(~ sum(!is.na(.)) > 1))  # Remove columns with all NAs or constant values

# Calculate the correlation matrix and tidy it
tidy_matrix <- cor(most_recent_correlation, use = "pairwise.complete.obs") %>%
  as_tibble(rownames = "company_1") %>%
  pivot_longer(-company_1, names_to = "company_2", values_to = "correlation") %>%
  mutate(
    company_1 = toupper(gsub("_close_return", "", company_1)),
    company_2 = toupper(gsub("_close_return", "", company_2))
  )

# Static heatmap
ggplot(tidy_matrix, aes(x = company_1, y = company_2, fill = correlation)) +
  geom_tile(color = "white") +
  scale_fill_distiller(palette = "RdBu", limits = c(-1, 1)) +
  labs(
    title = "Correlation Heatmap",
    x = "Companies",
    y = "Companies",
    fill = "Correlation"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.title = element_text(size = 12),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5)
  )
```

## Animated Monthly Correlation Heatmap:

```{r }
# Prepare monthly correlation matrices
tidy_correlation <- europe_stock_data %>%
  mutate(month_year = floor_date(date, "month")) %>%
  group_by(month_year) %>%
  reframe(across(ends_with("_close"), ~ (./lag(.) - 1) * 100, .names = "{.col}_return")) %>%
  nest(data = -month_year) %>%
  mutate(
    correlation_matrix = map(data, ~ {
      data_clean <- select(.x, ends_with("_return")) %>%
        select(where(~ sum(!is.na(.)) > 1))
      cor(data_clean, use = "pairwise.complete.obs")
    }),
    tidy_matrix = map(correlation_matrix, ~ as_tibble(.x, rownames = "company_1") %>%
      pivot_longer(-company_1, names_to = "company_2", values_to = "correlation"))
  ) %>%
  select(month_year, tidy_matrix) %>%
  unnest(tidy_matrix) %>%
  mutate(
    company_1   = toupper(gsub("_close_return", "", company_1)),
    company_2   = toupper(gsub("_close_return", "", company_2)),
    month_label = format(month_year, "%Y %B")  # "Year Month" label
  )

# Create the animated heatmap
p <- ggplot(tidy_correlation, aes(x = company_1, y = company_2, fill = correlation)) +
  geom_tile(color = "white") +
  scale_fill_distiller(palette = "RdBu", limits = c(-1, 1)) +
  labs(title = "Monthly Correlation Heatmap: {closest_state}", x = "", y = "", fill = "Correlation") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  transition_states(month_label, transition_length = 4, state_length = 2) + 
  ease_aes('linear')

# To render:
# animate(p, nframes = 150, fps = 10, width = 600, height = 500, renderer = gifski_renderer())
```

## Portfolio Analysis (Markowitz) Compute Daily Returns, Mean, and Covariance:

```{r }
# Calculate daily returns for all assets
daily_returns <- europe_stock_data %>%
  select(date, ends_with("_close")) %>%
  mutate(across(ends_with("_close"), ~ (./lag(.) - 1))) %>%
  filter(!is.na(rowMeans(select(., -date))))  # remove rows with all NAs

# Compute average daily returns and covariance matrix
mean_returns <- colMeans(daily_returns[-1], na.rm = TRUE)
cov_matrix   <- cov(daily_returns[-1], use = "pairwise.complete.obs")

# Annualize them (assuming ~252 trading days per year)
mean_returns_annualized <- mean_returns * 252
cov_matrix_annualized   <- cov_matrix * 252
```

## Generate Random Portfolios & Plot Efficient Frontier:

```{r }
set.seed(123)
num_portfolios <- 5000

portfolio_results <- replicate(num_portfolios, {
  # Generate random weights for 10 assets
  weights <- runif(10)
  weights <- weights / sum(weights)  # Sum to 1
  
  # Calculate portfolio return and risk
  expected_return <- sum(weights * mean_returns_annualized)
  portfolio_risk  <- sqrt(t(weights) %*% cov_matrix_annualized %*% weights)
  sharpe_ratio    <- expected_return / portfolio_risk
  
  c(expected_return, portfolio_risk, sharpe_ratio)
})

# Convert results to data frame
portfolio_results <- as.data.frame(t(portfolio_results))
colnames(portfolio_results) <- c("Return", "Risk", "Sharpe_Ratio")

# Plot the Efficient Frontier with LOESS curve
ggplot(portfolio_results, aes(x = Risk, y = Return)) +
  geom_point(aes(color = Sharpe_Ratio), alpha = 0.5) +
  geom_smooth(method = "loess", formula = y ~ x, color = "red", se = FALSE, linewidth = 1) +
  scale_color_gradient(low = "blue", high = "green") +
  labs(
    title = "Efficient Frontier (Annualized Returns with LOESS Curve)",
    x = "Portfolio Risk (Volatility, Annualized)",
    y = "Expected Return (Annualized)",
    color = "Sharpe Ratio"
  ) +
  theme_minimal()
```

## Interactive Efficient Frontier:

```{r }
# Fit LOESS model to the random portfolios
loess_fit <- loess(Return ~ Risk, data = portfolio_results, span = 0.5)

# Generate predicted values
risk_seq <- seq(min(portfolio_results$Risk), max(portfolio_results$Risk), length.out = 200)
smoothed_returns <- predict(loess_fit, newdata = data.frame(Risk = risk_seq))

# Create interactive plot
plot_ly() %>%
  add_markers(
    data = portfolio_results,
    x = ~Risk, y = ~Return, color = ~Sharpe_Ratio,
    type = 'scatter', mode = 'markers',
    colors = colorRamp(c("blue", "green")),
    marker = list(size = 6),
    name = "Portfolios"
  ) %>%
  add_lines(
    x = risk_seq, y = smoothed_returns, name = "LOESS Curve",
    line = list(color = 'red', width = 2)
  ) %>%
  layout(
    title = "Interactive Efficient Frontier with LOESS",
    xaxis = list(title = "Portfolio Risk (Volatility, Annualized)"),
    yaxis = list(title = "Expected Return (Annualized)"),
    legend = list(orientation = "h", x = 0.3, y = -0.2)
  )
```

```{r }
# Quadratic Programming to find optimal weights
Dmat <- 2 * cov_matrix_annualized
dvec <- rep(0, ncol(cov_matrix_annualized))

# Example constraints:
# 1) Sum of weights = 1
# 2) Each weight >= 0.05
# 3) Each weight <= 0.20 
Amat <- cbind(1, diag(1, ncol(cov_matrix_annualized)), -diag(1, ncol(cov_matrix_annualized)))
bvec <- c(1, rep(0.05, ncol(cov_matrix_annualized)), rep(-0.20, ncol(cov_matrix_annualized)))

result <- solve.QP(Dmat, dvec, Amat, bvec, meq = 1) 
optimal_weights <- result$solution

# Format the optimal weights
optimal_weights_df <- data.frame(
  Company        = toupper(gsub("_close", "", colnames(cov_matrix))),
  Optimal_Weight = round(optimal_weights * 100, 2)
)

optimal_weights_df

```

## Summary and Conclusion 
In this project, we walked through fetching data on Europe's top companies, plotting their locations, analyzing stock returns, and running a Markowitz optimization in R all in a few dozen lines of code.


## Publication 
https://medium.com/@karoly.takacs97/a-deep-dive-into-europes-top-10-stocks-with-r-1951f5c4e3c8
